Low Complexity Adaptive Sonar Imaging

### 2018-02-06 LCAforActiveSonarImaging/lca_submit2.tex
Major commit before messing with PhD thesis

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
% 
% \begin{itemize}
% \item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
% \item But:
% \begin{itemize}
% \item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
% \item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
% \item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix. LCA is only O($MW$). Even if Capon subspace/beamspace versions exist, they remain more complex than the LCA, and is not as ideally suited for implementation of SIMD hardware (GPUs, DSPs, ...).
% \end{itemize}
% \item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
% 
% The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions. For sonar arrays, this is achieved by applying weights to the array channels.
% 
% While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data. The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output. 
% 
We have studied the low complexity adaptive (LCA) beamformer in active sonar imaging. LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer. While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR seeks the optimal solution by estimating and inverting the array's spatial covariance matrix, while LCA selects the best performer from a predefined set of sensor windows.

We show that a robust MVDR implementation typically creates windows spanning between the rectangular and Hamming. We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each. We limit the steering to roughly half the -3\,dB width of the window's amplitude response. Using experimental data from the Kongsberg Maritime HISAS 1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS. On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR. This performance is achieved with a total of 6 windows; the rectangular and Kaiser window with $\beta=5$, in unsteered version, and left and right steered to the steering limit. Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference. Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.

@@

\IEEEPARstart{T}{he} best adaptive sonars are found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. Active human made sonars have little to no adaptability on transmit, but several methods have been investigated to adaptively post process the received data. Perhaps the best studied method for reconstructing a sonar image is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}. However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$). This is because MVDR relies on estimating and inverting a covariance matrix. In sonar systems with less than 32 channels the complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}. For larger systems the inversion step dominates the computation. This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}. Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}. However, the  mentioned methods are still relatively slow compared to DAS.

A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer. Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}. in clinical medical imaging~\cite{Synnevag2008}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams. LCA applies a set of predefined windows, or apodization weights, and selects the one offering the best sidelobe suppression. It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a reduced window solution space. Alternatively, it may be viewed as a multi-apodization technique. One such method is described by Stankwitz \emph{et al}. in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected. However, LCA is different in that it typically selects from a larger pool of windows and allows micro-steered variations of them. This gives LCA an adaptive freedom that is somewhere in between that of traditional multi-apodization techniques and MVDR.

Synnev\aa{}g \emph{et al}. suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions. In total he made LCA choose from 12 unique windows, 6 of which were micro-steered variants of a fairly wide Kaiser window, and with the rectangular window included. This was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration. While achieving good results, it was not made clear how how many windows LCA needs, or how they ideally should be steered. Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.

In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS 1030 synthetic aperture sonar (SAS), which operates at 100\,kHz with an element beamwidth of 23$^\circ$.

Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's -\,3dB resolution. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's -3\,dB width. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

@@

The output $z[n]$ of a beamformer is defined as the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{eq:beamformer_output}
\end{align}
%
where $w_m$ is the weight factor assigned to channel $m$. A weight set is commonly called a window, apodization or taper function. Applying a window that trails off towards the edges improves noise suppression at the cost of reduced resolution~\cite{Harris1978}. Real windows lead to symmetric responses, while complex weights allow asymmetric responses.

Conventional beamformers all have static weights. The reference method is the delay-and-sum (DAS) beamformer, also know as the backprojection algorithm. It delays each pixel into focus, then applies a suitable window, and finally sums the data. The virtue of DAS is its simplicity, robustness to parameter errors, linear processing of the image and the ease of which it can be implemented on parallel hardware.

Adaptive beamformers are dynamic and seek to adjust the array response to better fit the incoming wavefield. This may be achieved by allowing either the weights or delays to change, or both. One of the most extensively studied of these methods is the minimum variance distortion-less response (MVDR) beamformer.

@@

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/calc_kaiser_3dB.eps}%
\else%
\includegraphics[width=\linewidth]{gfx/calc_kaiser_3dB.eps}%
\fi%
\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\text{-3dB}}(\beta)]$. The angle $\phi_{\text{-3dB}}(\beta)$ is the amount of steering needed for the steered response to have a -3dB crossing that is exactly half that of the unsteered window. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
\end{figure}

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/cross_tagged.png}%
\else%
\includegraphics[width=\linewidth]{gfx/cross_tagged.png}%
\fi%
\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m resolution test cross attached to an anchor with a diameter of approximately 13\,cm. Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
\end{figure}

@@

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/mvdr_selected_windows_holmengraa.pdf}%
\caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
\emph{Left:}\hfill
\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with the pixel groups are indicated with red boxes.}\protect\\\hspace{\textwidth}
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups in the shadow, highlight and speckle region of the image. The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale. The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight. The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
\end{figure*}

\begin{figure*}[t]%
\includegraphics[width=\textwidth]{gfx/oversampling_mosaic_bounds.pdf}%
\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied. The upper left image is equal to DAS with a rectangular window. For all images we measure the lateral distance $\Delta x$ between the two -3dB points of the anchor (red object). These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
\newline
\emph{Upper bound $\beta$:}\hfill
\parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$. \Fig{mvdr_selected_windows} suggests that MVDR tend to choose windows with levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
\emph{Upper bound $\phi$:}\hfill
\parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
}\label{oversampling_mosaic_bounds}
\end{figure*}

@@

Since pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
%
\begin{align}
s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function. However, we observed no visual improvement doing this, and the effect of micro-steered windows is hard to predict. Henceforth no lateral averaging will be applied.
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

Each window's response is constrained to unit gain in the look direction. Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}. To limit this we place an upper bound to the steering, chosen such that wide windows is allowed to be steered more than narrow ones. We call this upper bound for steering $\phi_{\text{3dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window. Using a numerical search for the steering angle $\phi$ that fulfilled this criterion for a wide range of $\beta$-values, we obtained a third degree polynomial fit to these points:
%
\begin{align}
\phi_\text{3dB}(\beta) \approx \minus{}0.002894\beta^3 + 0.02451\beta^2 - 0.01308\beta + 0.2980\label{eq:steering_kaiser_3dB}
\end{align}
%
Henceforth any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

@@

Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers. The images appear more visually pleasing and detailed up to 8. We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale. In \mediaI{} we visualize the effect of changing the lateral oversampling factor on a set of LCA images. Upon display the images are all bilinearly upinterpolated to the same size. 
% 
% \todo{Need $M$ in the image sector. Why???}

@@

To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}. It is a high resolution synthetic aperture sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ receiver element beamwidth.

Two different scenes will be studied. One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}. It measures 68\;m by 9\;m and fills most of the highlighted sector. This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime. The other scene is of an 1\,m by 1\,m iron resolution test cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}). This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71). Both scenes are imaged in sectorscan mode.

@@

Unlike the method described by Synnev\aa{}g~\cite{Synnevag2008} we do not let LCA choose from inverted Kaiser windows. These have a mainlobe width narrower than that of the rectangular window, but at the expense of a very poor white noise gain. We have found that these windows hardly ever get used on experimental data, in particular steered versions of them. We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of \minus{}13\,dB.

\mediaII{} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$. At $L=1$ the window responses are rectangular in all areas in the image. At $L=2$ we observe window responses with slight amplitude variations in the highlight region, but with the phase being 0$^\circ$ in the illuminated seafloor sector. Already at $L=2$ the MVDR is able to greatly suppress noise. This is a common observation; adding a little flexibility to adapt to the scene has a dramatic effect, but further allowing full flexibility is much less significant. The media file also shows that MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.

@@

% \begin{table}[!b]\centering%\normalsize
% \begin{tabular}[c]{l c c c}\hline
% \rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
% Symmetry             & Moderate  & High   & High \\
% Mainlobe width       & Narrow    & Wide   & Moderate \\
% Sidelobe suppression & Low       & High   & Moderate \\
% Steering             & Low       & Up to -3dB width & Moderate \\
% Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
% \end{tabular}
% \caption{Summary}\label{tab:summary}
% \end{table}%

LCA seeks to improve image resolution and contrast by minimizing the power of noise and interference in each image pixel. This is the same optimization criterion as MVDR uses, but instead of computing an optimal array window like MVDR does, it selects the best window out of a predefined set. Hence, it can be viewed as either a reduced and discrete window space version of MVDR, or as an advanced multi-apodization technique. What we have studied in this paper is which windows we should ideally let LCA choose from.

>>>>

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
% 
% \begin{itemize}
% \item Dynamic weights $\Rightarrow$ greater contrast/resolution potential
% \item But:
% \begin{itemize}
% \item Active system $\Rightarrow$ noise and signal correlated $\Rightarrow$ adaptive beamformers no longer robust.
% \item Subarray averaging decorrelate the noise, but in the process we sacrifice resolution.
% \item Computationally complex O($M^3$) due to inversion of a spatial covariance matrix. LCA is only O($MW$). Even if Capon subspace/beamspace versions exist, they remain more complex than the LCA, and is not as ideally suited for implementation of SIMD hardware (GPUs, DSPs, ...).
% \end{itemize}
% \item LCA is inherently robust, is ``parameter free'' once setup with a well designed window set, and can often be implemented with minor modifications to already existing DAS beamformers.
% 
% The angular resolution and contrast in active sonar images depend on the beamformer's ability to receive signals from directions of interest, while suppressing noise and interference emanating from other directions. For sonar arrays, this is achieved by applying weights to the array channels.
% 
% While classical beamformers use predefined windows, adaptive beamformers estimate the optimal window by analytical evaluation of the data. The minimum variance (MV) beamformer, for instance, calculates the set of weights that minimises the variance of the beamformer's output. 
% 
We have studied the low complexity adaptive (LCA) beamformer in active sonar imaging. LCA can be viewed as either a simplification of the minimum variance distortionless response (MVDR) beamformer, or as an adaptive extension to the delay and sum (DAS) beamformer. While both LCA and MVDR attempt to minimize the power of noise and interference in the image, MVDR achieves this by computing optimal array weights from the spatial statistics of the wavefield, while LCA selects the best performing weights out of a predefined set.

To build confidence in the LCA method we show that a robust MVDR implementation typically creates weights spanning between the rectangular and Hamming window function. We let LCA select from a set of Kaiser windows with responses in this span, and add some steered variations of each. We limit the steering to roughly half the \minus{}3\,dB width of the window's amplitude response. Using experimental data from the Kongsberg Maritime HISAS1030 sonar we find that LCA and MVDR produce nearly identical images of large scenes, both being superior to DAS. On point targets LCA is able to double the resolution compared to DAS, or provide half that of MVDR. This performance is achieved with a total of 6 windows; the rectangular window and the Kaiser window with $\beta=5$, in an unsteered version, and versions that are left and right steered to the steering limit. Slightly smoother images are produced if the window count is increased to 15, but past this we observe minimal difference. Finally we show that LCA works just as well if Kaiser windows are substituted with trigonometric ones.

@@

% The promise is already demonstrated in nature.
% 
% The sonar systems of the future 
% 
% Future sonar systems 
% 
% Great promise in adaptivity
% 
% The greatest untapped 
% 
% The sonars of the future might benefit greatly from adaptivity.
% 
% 
% Future sonars will likely be

% To make a sonar truly perform better, we need to make it smarter. 
% 
% \IEEEPARstart{F}{uture} sonar systems will merit a much more sophisticated level of adaptivity. 
% 
% \IEEEPARstart{T}{he} next generation sonar systems will be adaptive.
% 
% Making sonars adaptive is the way to take them into the future. 
% 
% The virtue of future sonar systems will be sophisticated adaptivity. 
% 
% Intelligent
% 
% Reactive
% 
% Observe, process, adapt, act. 
% 
% Adaptivity - sense, process, respond
% 
% Faster, smarter
% 
% The best sonars in nature are adaptive. 
% Nature allures us to be this bold because

\IEEEPARstart{T}{he} best sonars in existence is perhaps found in nature. Bats, for instance, use a high frequency sonar to detect, identify and track prey with amazing precision in caves amongst a myriad of other bats. No man made sonar can currenly match this feat, much due to the challenge of processing and adapting to such a complex environment fast enough. This is why human made sonars have little to no adaptivity on transmit. However, several methods have been investigated to adaptively post process the received data. Perhaps the best studied method for reconstructing a sonar image out of the data is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It adapts the sonar array's spatial response to minimize the influence of noise and interference in the final image.

In many cases MVDR can improve the contrast and resolution of active sonar images compared to conventional static methods~\cite{Blomberg2013,Blomberg2012a,Lo2004}. However, while the computational complexity of conventional beamformers are linear with the number of channels, O($M$), MVDR is at O($M^3$). This is because MVDR relies on estimating and inverting a covariance matrix. The estimation step dominates the computation in sonar systems with less than 32 channels, but here the computational complexity can be reduced and the implementation accelerated significantly using graphics computing units (GPUs)~\cite{Buskenes2014}. For larger systems the inversion step dominates. This can be dealt with by approximating the full covariance matrix with a small one using space reduction techniques such as beamspace processing~\cite{Asen2013}, or the closely related principal component analysis (PCA) method~\cite{Kim2014}. Another alternative is to assume spatial stationarity to form more easily invertible Toepliz matrices~\cite{Asl2012}. However, the mentioned methods are still relatively slow compared to DAS.

A much faster and simpler alternative to MVDR is the Low Complexity Adaptive (LCA) beamformer. Based on an idea by Vignon~\cite{Vignon2008}, it was first introduced by Synnev\aa{}g \emph{et al}. in clinical medical ultrasound imaging~\cite{Synnevag2011}, who demonstrated its ability to obtain very similar image quality to MVDR in systems with focused transmit beams. LCA applies a set of predefined and static windows, or apodization weights, and selects the one offering the best noise suppression. It relies on the same optimization criterion as MVDR, thus may be considered a version of MVDR with a discrete and static window solution space. Alternatively, it may be viewed as a multi-apodization technique. One such method is described by Stankwitz \emph{et al}. in radar imaging~\cite{Stankwitz1995}, where the best out of 2 or 3 windows are selected. However, LCA is different in that it typically selects from a larger pool of windows and allows phase-steered variations of them. This gives LCA an adaptive freedom that lies somewhere in between that of traditional multi-apodization techniques and MVDR. How similarly LCA performs compared to MVDR depends on how well the predefined windows represent the constrained solution space of MVDR.

% While using weights that span a bigger space gives LCA greater freedom, not The bigger the space the weights span, the greater the adaptive freedom, but only the solutions that are valid in the constrained solution space can make a difference.

% Instead of adaptively computing the weights, it adaptively selects them.

Synnev\aa{}g \emph{et al}. suggested using a window set comprised of rectangular, Kaiser (or Kaiser-Harris) and inverted Kaiser functions. In total he made LCA choose from 12 unique windows, 6 of which were phase-shifted versions of a fairly wide Kaiser window, and one were rectangular. Phase-shifting the windows has the effect of slighly adjusting the angle the array is steered towards. Synnev\aa{}g's composition of windows was motivated by the desire to use wide windows with maximum sidelobe suppression near the receiver, and narrow windows at greater depths for maximum sensitivity and depth penetration. While achieving good results, it was not made clear how many windows LCA needs, or how they ideally should be steered. Also, it is not apparent how LCA performs in sonar systems where the transmit beam is unfocused.

In this study we expand our previous work of applying LCA to active sonar imaging~\cite{Buskenes2011}. We use variations of the parametric Kaiser window, but also compare it to the trigonometric window function. We extend Synnev\aa{}g's work by steering all our windows and not just some, and demonstrate how LCA's performance is affected by the type and number of windows included. Our results are obtained using experimental data from the 32 element Kongsberg Maritime HISAS1030 sonar, which operates at 100\,kHz with an \minus{}3\,dB element opening angle of 23$^\circ$. Although this system was designed as synthetic aperture sonar (SAS) where an image is synthesized from several pings of data, we will only create images out of single pings (sectorscan imaging) in this study.

Our results show that when MVDR is modified to work in active sonar imaging, it computes weights with spatial responses that tend to be symmetric and steered within a fraction of the window's \minus{}3\,dB resolution. When letting LCA choose from Kaiser windows with similar responses, we obtain images with improved noise suppression and resolution. The resolution gain is made predictable by limiting the maximum steering to a fraction of the window's \minus{}3\,dB width. Overall, LCA produces images that are similar to MVDR, but with a resolution in between that of DAS and MVDR. LCA is also inherently robust, easy to implement, and fairly easy to understand.

@@

The output $z[n]$ of a beamformer is defined as the weighted sum of all the delayed data samples:
%
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{eq:beamformer_output}
\end{align}
%
where $w_m$ is the weight factor assigned to channel $m$. A weight set is commonly called a window, apodization or taper function. When the window is real its response is symmetric. Applying a real window that trails off towards the edges creates a response with lower sidelobe levels and wider mainlobe, which translates into improved noise suppression at the cost of reduced resolution, respectively~\cite{Harris1978}.

Conventional beamformers all have static and usually real weights. The reference method is the delay-and-sum (DAS) beamformer, also know as the backprojection algorithm. It delays each pixel into focus, then applies a suitable window, and finally sums the data. The virtue of DAS is its simplicity, robustness to parameter errors, linear processing of the image and the ease of which it can be implemented in parallel hardware.

Adaptive beamformers are dynamic and seek to adjust the array response to better fit the incoming wavefield. This may be achieved by allowing either the weights or delays to change, or both. Here the weights are usually complex, which allows asymmetric responses. One of the most extensively studied adaptive methods is the minimum variance distortion-less response (MVDR) beamformer.

@@

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/buske1_online.eps}%
\else%
\includegraphics[width=\linewidth]{gfx/buske1_online.eps}%
\fi%
\caption{Each Kaiser window is steered in the interval $\phi=[0, \phi_{\mathrm{3\,dB}}(\beta)]$. The angle $\phi_{\mathrm{3\,dB}}(\beta)$ is the amount of steering needed for the steered response to have a \protect\minus{}3\,dB crossing that is exactly half that of the unsteered window. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.}\label{windows_steering}
\end{figure}

\begin{figure}[tbp]\centering%
\ifPeerReview%
\includegraphics[width=.5\linewidth]{gfx/buske2_online.png}%
\else%
\includegraphics[width=\linewidth]{gfx/buske2_online.png}%
\fi%
\caption{One of the objects used to test the beamformers is a 1\,m by 1\,m resolution test cross attached to an anchor with a diameter of approximately 13\,cm. Source image curtesy of Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).}\label{cross}
\end{figure}

@@

\begin{figure*}[t]\centering%
\includegraphics[width=\linewidth]{gfx/buske3_online.pdf}%
\caption{\emph{Determining LCA window type}: MVDR image with typical frequency responses for windows used in various pixel regions.\newline
\emph{Left:}\hfill
\parbox[t]{.95\linewidth}{MVDR sectorscan image of the oiltanker Holmengraa, with the pixel groups are indicated with red boxes.}\protect\\\hspace{\textwidth}
\emph{Right:}\hfill
\parbox[t]{.95\linewidth}{MVDR window amplitude and phase responses computed from 40x40 pixel groups for the shadow, highlight and speckle region of the image. The responses are overlayed each other and the amount of overlap is colored using a logarithmic scale. The size of the pixel groups were chosen ad-hoc for the histograms to be near invariant to a shift in position within the same region. Since each pixel is pre-delayed into focus the unsteered responses all have their center at broadside. The dashed red lines at -13\,dB and -43\,dB marks the peak sidelobe levels of an unsteered rectangular and Hamming window, respectively. Note how the responses are more or less symmetric, with very little steering in the shadow, moderate steering in speckle and steering within roughly 3\,dB in highlight. The phase varies most in the highlight region where we see the highest contrast.} }\label{mvdr_selected_windows}
\end{figure*}

\begin{figure*}[t]%
\includegraphics[width=\textwidth]{gfx/buske4_online.pdf}%
\caption{\emph{Determining Kaiser parameter boundaries}: LCA images of the resolution test cross and anchor created with a large and equal size window database, but where the upper bounds for $\beta$ and $\phi$ is varied. The upper left image is equal to DAS with a rectangular window. For all images we measure the lateral distance $\Delta x$ between the two \protect\minus{}3\,dB points of the anchor (red object). These are specified relatively to the reference distance  $\Delta x_\text{ref}$ from the upper left DAS image.
\newline
\emph{Upper bound $\beta$:}\hfill
\parbox[t]{.89\linewidth}{Observe that LCA gets better at suppressing sidelobes in the image as we increase the upper bounds for $\beta$. \Fig{mvdr_selected_windows} suggests that MVDR tend to choose windows with levels lower than that of Kaiser with $\beta=2$, but here we observe further improvement going to $\beta=5$.}\newline
\emph{Upper bound $\phi$:}\hfill
\parbox[t]{.89\linewidth}{As we increase the upper bound of the steering $\phi$, we also increase the lateral image resolution. }
}\label{oversampling_mosaic_bounds}
\end{figure*}

@@

If the pixels are correlated laterally, we can include a weighted combination of these as well to improve the variance estimation:
%
\begin{align}
s_z^2 = \frac{1}{N_x N_k} \sumb{\nmat{\scriptstyle x'=\\[-2mm]\scriptstyle x-X}}{x+X} \sumb{\nmat{\scriptstyle n'=\\[-2mm]\scriptstyle n-K}}{n+K} \omega[x',n']\big| z[x',n'] \big|^2
\end{align}
%
where $N_x = 2X+1$ is the number of azimuth lines to average over, and $\omega[x',n']$ is a normalized 2 dimensional weight function. Since we oversample slightly laterally when we delay each pixel (see Section \ref{sec:lca_oversampling}), we thought the lateral correlation to be sufficient to benefit from this. However, we observed no visual improvement from applying this technique, and decided not to use it.
% 
% \begin{itemize}
% \item \emph{Window function}
% \item \emph{Parameter boundaries}
% \item \emph{Number of windows needed}
% \item \emph{Oversampling need}
% \item \emph{Image quality}
% \end{itemize}

@@

Each window's response is constrained to unit gain in the look direction. Hence, when it is steered the white noise gain must increase and the the signal-to-noise ratio decrease, as shown in \Fig{windows_steering}. To limit this we place an upper bound to the steering, chosen such that wide windows are allowed to be steered more than narrow ones. We call this upper bound for steering $\phi_{\mathrm{3\,dB}}(\beta)$, and define it as the steering angle needed for the steered response to have a \minus{}3\,dB crossing that is exactly half that of the unsteered window. The method we devised for finding it is described in Appendix \ref{appendix:steering_angles}.

Henceforth any steering will be specified relative to this upper bound. With each window steered this way we expect the resolution gain to remain predictable and independent of $\beta$, and we also effectively constrain the white noise gain of the beamformer.

@@

Since LCA operates with the same optimization criterion as MVDR, we will be using approximately a factor 8 lateral oversampling for both beamformers. The images appear more visually pleasing and detailed up to 8. We consider an absolute minimum to be 2, due to the non-linear nature of delaying pixels that isn't strictly in the far field, and of displaying the absolute value of the pixels on a decibel scale. In \mediaI{}\footnote{For media location refer to Appendix \ref{app:source_code}.} we visualize the effect of changing the lateral oversampling factor on a set of LCA images. Upon display the images are all bilinearly upinterpolated to the same size. 
% 
% \todo{Need $M$ in the image sector. Why???}

@@

To test the performance of the MVDR and LCA beamformers, we have processed data acquired by the Kongsberg Maritime HUGIN AUV carrying their 32 element HISAS1030 sonar~\cite{Hansen2011}. It is a high resolution sonar with 1.2\;m array length, 100\;kHz operating frequency, 30\;kHz bandwidth and 23$^\circ$ element \minus{}3\,dB opening angle.

Two different scenes will be studied. One of the 1500 DWT oil tanker wreck Holmengraa lying at a slanted seabed at 77\;m depth outside of Horten, Norway~\cite{holmengraa}. It measures 68\;m by 9\;m and fills most of the highlighted sector. This data was collected by the Norwegian Defense Research Establishment and Kongsberg Maritime. The other scene is of an 1\,m by 1\,m iron resolution test cross, connected to a an anchor with 13\,cm diameter (\Fig{cross}). This data was collected by the Bundeswehr Technical Center for Ships and Naval Weapons, Maritime Technology and Research (WTD 71).

@@

Unlike the method described by Synnev\aa{}g~\cite{Synnevag2011} we do not let LCA choose from inverted Kaiser windows. These have a mainlobe width narrower than that of the rectangular window, but at the expense of a very poor white noise gain. We have found that these windows hardly ever get used on experimental data, in particular steered versions of them. We can infer the same from \Fig{mvdr_selected_windows} by noting that the maximum sidelobe level rarely exceeds the rectangular window level of \minus{}13\,dB.

\mediaII{}\footnote{For media location refer to Appendix \ref{app:source_code}.} animates how \Fig{mvdr_selected_windows} changes as a function of the MVDR subarray length $L$ and temporal averaging $K$. At $L=1$ the window responses are rectangular in all areas in the image. At $L=2$ we observe window responses with slight amplitude variations in the highlight region, but with the phase being 0$^\circ$ in the illuminated seafloor sector. Already at $L=2$ the MVDR is able to greatly suppress noise. This is a common observation; adding a little flexibility to adapt to the scene has a dramatic effect, but further allowing full flexibility is much less significant. The media file also shows that MVDR can be run with subarray sizes $L\in[M/2,5M/8]$, but only with temporal averaging $K=1$ or above.

@@

% \begin{table}[!b]\centering%\normalsize
% \begin{tabular}[c]{l c c c}\hline
% \rowcolor{tabBlue} & \bf Highlight & \bf Shadow & \bf Speckle  \\\hline
% Symmetry             & Moderate  & High   & High \\
% Mainlobe width       & Narrow    & Wide   & Moderate \\
% Sidelobe suppression & Low       & High   & Moderate \\
% Steering             & Low       & Up to \minus{}3\,dB width & Moderate \\
% Oversampling         & 2x-8x     & 2x-8x  & 2x-8x
% \end{tabular}
% \caption{Summary}\label{tab:summary}
% \end{table}%

LCA seeks to improve image resolution and contrast by minimizing the power of noise and interference in each image pixel. This is the same optimization criterion as MVDR uses, but instead of computing an optimal array window like MVDR does, it selects the best window out of a predefined set. Hence, it can be viewed as either a reduced and discrete window space version of MVDR, or as an advanced multi-apodization technique. In this paper we have studied which windows we should ideally let LCA choose from.
